A comparison of design-based and model-based approaches for spatial data.

### 2021-06-11 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.Rmd
minor revisions and comments.

---
title: A comparison of design-based and model-based approaches for spatial data.
author:
  - name: Michael Dumelle
    email: Dumelle.Michael@epa.gov
    affiliation: USEPA
    footnote: 1
  - name: Matthew Higham
    email: mhigham@stlaw.edu
    affiliation: STLAW
    footnote: 1
  - name: Lisa Madsen
    affiliation: OSU
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Jay M. Ver Hoef
    affiliation: NOAA
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Math, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  This is the abstract.

@@

There are two general approaches for using data to make statistical inferences about a population: design-based approaches and model-based approaches. When data cannot be obtained for all units in a population (known as population units), data on a subset of the population units is collected in a  sample. In the design-based approach, inferences about the underlying population are informed from a probabilistic process in which population units are selected to be in the sample. Alternatively, in the model-based approach, inferences are made from specific assumptions made about the underlying process that generated the data. Each paradigm has a deep historical context [@sterba2009alternative] and its own set of general advantages [@hansen1983evaluation].

@@

The model-based approach assumes the data are a random realization of a process. Randomness is often incorporated through distributional assumptions on the data-generating process. Instead of estimating fixed but unknown parameters (as in the design-based approach), the goal of model-based inference in the spatial context is often \emph{prediction} of an unknown quantity. For example, suppose the realized mean of all population units is the quantity of interest. Instead of \emph{estimating} a fixed unknown mean, we are \emph{predicting} the value of the mean, a random variable. We know that if we sampled all population units, we would have an exact prediction for the mean of our one realized process, without any uncertainty. But the true mean of the spatial process that generated our realized data is still not known. When predicting the realized mean, we typically are not interested in the underlying process' true mean.

Assuming the data is a realization of a specific data-generating process yields predictors that are linked to distributional assumptions. These distributional assumptions are used to derive prediction intervals. The distributional assumptions allow the prediction intervals can be more precise. @cressie1993statistics and @schabenberger2017statistical provide reviews of model-based approaches for spatial data.

@@

Sampling designs can incorporate spatial locations to obtain samples that are spatially balanced with respect to the population [@stevens2004spatially]. A sample is spatially with respect to the population if the sampled population units are a miniature of the population units. A sample is a miniature of the population if the distribution of the sampled population units mirrors the density of all population units. Spatial balance with respect to the population is is different than spatial balance with respect to geography. A sample that is spatially balanced with respect to geography is spread out in some type of equidistant manner over geographical space and are not meant to be miniatures of the population. When we refer to spatial balance henceforth, we mean spatial balance with respect to the population.

Spatially balanced samples are useful because they tend to yield estimates that have lower variance than estimates constructed from sampling designs lacking spatial balance [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @benedetti2017spatiallyreview; @wang2013design]. To quantify spatial balance, @stevens2004spatially proposed loss functions based on Voroni polygons. The first spatiallly balanced sampling algorithm that saw widespread use was the Generalized Random Tessellation Stratified [@stevens2004spatially]. Since GRTS was developed, several other spatially balanced sampling algorithms have emerged, including the Local Pivotal Method [@grafstrom2012spatially; @grafstrom2018spatially], Spatially Correlated Poisson Sampling [@grafstrom2012spatiallypoisson], Balanced Acceptance Sampling [@robertson2013bas], Within-Sample-Distance [@benedetti2017spatially], and Halton Iterative Partitioning [@robertson2018halton] algorithms. We focus on the Generalized Random Tessellation Stratified (GRTS) algorithm to select spatially balanced sampling because the algorithm has several attractive properties detailed by @stevens2004spatially and @dumelle2021spsurvey.

@@

After collecting a sample, the data are used to estimate population parameters. The Horvitz-Thompson estimator [@horvitz1952generalization] and its continuous analog [@cordy1993extension] yield unbiased estimates of population means and totals. If $\tau$ is a population total, then the Horvitz-Thompson estimator of $\tau$ (denoted by $\hat{\tau}_{ht}$), is given by
\begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align}
where $Z_i$ and $pi_i$ are the observed value and inclusion probability of the $i$th population unit selected in the sample. @horvitz1952generalization and @sen1953estimate provide variance estimators for $\hat{\tau}_{ht}$, but they have two drawbacks. First, they rely on calculating $\pi_{ij}$, the probability that population unit $i$ and population unit $j$ are included in the sample, which can be very difficult to calculate. Second, they ignore the spatial locations of the population units. To address these drawbacks, @stevens2003variance proposed a local neighborhood variance estimator. The local neighborhood variance estimator does not rely on $\pi_{ij}$, and it incorporates spatial locations by assigning higher weights to nearby observations. @stevens2003variance show this variance estimators tends to reduce the variability associated with estimating $\tau$. This yields confidence intervals for $\tau$ that are narrower than confidence intervals constructed from variance estimators ignoring spatial locations.

@@

Finite Population Block Kriging (FPBK) is an alternative to samipling-based methods [@verhoef2008spatial]. FPBK expands the geostatistical kriging framework to the finite population setting. Instead of basing inference off of a specific sampling design, we assume the data were generated by a spatial process with parameters that can be estimated using the framework of a model.

>>>>

---
title: A comparison of design-based and model-based approaches for spatial data.
author:
  - name: Michael Dumelle
    email: Dumelle.Michael@epa.gov
    affiliation: USEPA
    footnote: 1
  - name: Matt Higham
    email: mhigham@stlaw.edu
    affiliation: STLAW
    footnote: 1
  - name: Lisa Madsen
    affiliation: OSU
  - name: Anthony R. Olsen
    affiliation: USEPA
  - name: Jay M. Ver Hoef
    affiliation: NOAA
address:
  - code: USEPA
    address: United States Environmental Protection Agency, 200 SW 35th St, Corvallis, Oregon, 97333
  - code: STLAW
    address: Saint Lawrence University Department of Math, Computer Science, and Statistics, 23 Romoda Drive, Canton, New York, 13617
  - code: OSU
    address: Oregon State University Department of Statistics, 239 Weniger Hall, Corvallis, Oregon, 97331
  - code: NOAA
    address: Marine Mammal Laboratory, Alaska Fisheries Science Center, National Oceanic and Atmospheric Administration, Seattle, Washington, 98115
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  This is the abstract.

@@

There are two general approaches for using data to make statistical inferences about a population: design-based approaches and model-based approaches. When data cannot be obtained for all units in a population (known as population units), data on a subset of the population units is collected in a  sample. In the design-based approach, inferences about the underlying population are informed from a probabilistic process in which population units are selected to be in the sample. Alternatively, in the model-based approach, inferences are made from specific assumptions about the underlying process that generated the data. Each paradigm has a deep historical context [@sterba2009alternative] and its own set of general advantages [@hansen1983evaluation].

@@

The model-based approach assumes the data are a random realization of a process. Randomness is often incorporated through distributional assumptions on the data-generating process. Instead of estimating fixed but unknown parameters (as in the design-based approach), the goal of model-based inference in the spatial context is often \emph{prediction} of an unknown quantity. For example, suppose the realized mean of all population units is the quantity of interest. Instead of \emph{estimating} a fixed unknown mean, we are \emph{predicting} the value of the mean, a random variable. We know that if we sampled all population units, we would have an exact prediction for the mean of our one realized process, without any uncertainty. But the true mean of the spatial process that generated our realized data is still not known. When predicting the realized mean, we typically are not interested in the underlying process's true mean.
<!-- (and is often not of interest) -->

Assuming the data is a realization of a specific data-generating process yields predictors that are linked to distributional assumptions. These distributional assumptions are used to derive prediction intervals. The distributional assumptions allow the prediction intervals to be more precise. @cressie1993statistics and @schabenberger2017statistical provide reviews of model-based approaches for spatial data.

@@

<!-- a sample is spatially balanced -->

Sampling designs can incorporate spatial locations to obtain samples that are spatially balanced with respect to the population [@stevens2004spatially]. A sample is spatially with respect to the population if the sampled population units are a miniature of the population units. A sample is a miniature of the population if the distribution of the sampled population units mirrors the density of all population units. Spatial balance with respect to the population is different than spatial balance with respect to geography. A sample that is spatially balanced with respect to geography is spread out in some type of equidistant manner over geographical space and is not meant to be miniatures of the population. When we refer to spatial balance henceforth, we mean spatial balance with respect to the population.

<!-- shorten? Spatial balance with respect to the population is different than spatial balance with respect to geography. A sample that is spatially balanced with respect to geography is spread out in some type of equidistant manner over geographical space and is not meant to be miniatures of the population. When we refer to spatial balance henceforth, we mean spatial balance with respect to the population.  -->

Spatially balanced samples are useful because they tend to yield estimates that have lower variance than estimates constructed from sampling designs lacking spatial balance [@stevens2004spatially; @barabesi2011sampling; @grafstrom2013well; @robertson2013bas; @benedetti2017spatiallyreview; @wang2013design]. To quantify spatial balance, @stevens2004spatially proposed loss functions based on Voroni polygons. The first spatially balanced sampling algorithm that saw widespread use was the Generalized Random Tessellation Stratified [@stevens2004spatially]. Since GRTS was developed, several other spatially balanced sampling algorithms have emerged, including the Local Pivotal Method [@grafstrom2012spatially; @grafstrom2018spatially], Spatially Correlated Poisson Sampling [@grafstrom2012spatiallypoisson], Balanced Acceptance Sampling [@robertson2013bas], Within-Sample-Distance [@benedetti2017spatially], and Halton Iterative Partitioning [@robertson2018halton] algorithms. We focus on the Generalized Random Tessellation Stratified (GRTS) algorithm to select spatially balanced sampling because the algorithm has several attractive properties detailed by @stevens2004spatially and @dumelle2021spsurvey.

<!-- algorithms. -->

@@

<!-- This makes it seem like HT is being used to estimate population parameters in GRTS? Needs to be in there, but maybe a sentence making it clear that this isn't being used? Or just change the order and talk about Local variance first and then mention these for more general settings? -->

After collecting a sample, the data are used to estimate population parameters. The Horvitz-Thompson estimator [@horvitz1952generalization] and its continuous analog [@cordy1993extension] yield unbiased estimates of population means and totals. For example, if $\tau$ is a population total, then the Horvitz-Thompson estimator of $\tau$ (denoted by $\hat{\tau}_{ht}$), is given by
\begin{align}\label{eq:ht}
  \hat{\tau}_{ht} = \sum_{i = 1}^n Z_i \pi_i^{-1},
\end{align}
where $Z_i$ and $\pi_i$ are the observed value and inclusion probability of the $i$th population unit selected in the sample. @horvitz1952generalization and @sen1953estimate provide variance estimators for $\hat{\tau}_{ht}$, but they have two drawbacks. First, they rely on calculating $\pi_{ij}$, the probability that population unit $i$ and population unit $j$ are included in the sample, which can be very difficult to calculate. Second, they ignore the spatial locations of the population units. To address these drawbacks, @stevens2003variance proposed a local neighborhood variance estimator. The local neighborhood variance estimator does not rely on $\pi_{ij}$, and it incorporates spatial locations by assigning higher weights to nearby observations. @stevens2003variance show this variance estimators tends to reduce the variability associated with estimating $\tau$. This yields confidence intervals for $\tau$ that are narrower than confidence intervals constructed from variance estimators ignoring spatial locations.

@@

Finite Population Block Kriging (FPBK) is an alternative to sampling-based methods [@verhoef2008spatial]. FPBK expands the geostatistical kriging framework to the finite population setting. Instead of basing inference off of a specific sampling design, we assume the data were generated by a spatial process with parameters that can be estimated using the framework of a model.
